---
title: "RichR"
subtitle: "Enrichment for Diseases in a Set of Genes"
excerpt: "It provides a list of genes associated to diseases (g2d$clean and g2d$complete) based on the following  4 publications (GS2D, Fontaine (2016) <doi:10.18547/gcb.2016.vol2.iss1.e33>, DisGeNET, Pinero (2016) <doi:10.1093/nar/gkw943> Berto2016, Berto (2016) <doi:10.3389/fgene.2016.00031> and  PsyGeNET, Sacristan (2015) <doi:10.1093/bioinformatics/btv301>). Those lists were combined and manually curated to have matching disease names.  When provided a list of gene names, it calculates the disease enrichment of the gene set. The enrichment is calculated using proportion test and Fisher's exact test. Adjusted fdr p-values are returned alongside with p-values combined using the Fisher's method."
date: 2019-02-13
date_end: "2019-02-13"
author: "Deisy Morselli Gysi"
location: "CRAN"
draft: false
# layout options: single, single-sidebar
layout: single
categories:
- r packages
- gene enrichment
- disease association
links:
- icon: box-open
  icon_pack: fas
  name: package
  url: https://cran.r-project.org/web/packages/RichR/
---



It provides a list of genes associated to diseases (`g2d$clean` and `g2d$complete`) based on the following  4 publications (GS2D, Fontaine (2016) [doi:10.18547/gcb.2016.vol2.iss1.e33], DisGeNET, Pinero (2016) [doi:10.1093/nar/gkw943] Berto2016, Berto (2016) [doi:10.3389/fgene.2016.00031] and  PsyGeNET, Sacristan (2015) [doi:10.1093/bioinformatics/btv301]). Those lists were combined and manually curated to have matching disease names.  When provided a list of gene names, it calculates the disease enrichment of the gene set. The enrichment is calculated using proportion test and Fisher's exact test. Adjusted fdr p-values are returned alongside with p-values combined using the Fisher's method. 

You can download the package from CRAN using:

``install.packages('RichR')``


## Input data


The input data for RichR are: the Background, that is the list of genes to be used as background control, the Genes2Dis, a data.frame containing genes and association to phenotypes and disorders and Genes, a vector of genes that should be tested for enrichment. 

We recommend using `g2d$clean` as Genes2Dis. This is a manually curated list of genes and association to disorders.


Usage

```{r}
require(RichR)
data('g2d')

g2d_clean = g2d$clean
```



The user can choose a particular disorder, or use the whole disease set to compare to

```{r}
g2d_ASD = subset(g2d_clean, 
                 g2d_clean$Disease %in% c('AUTISM'))

Enrichment(Background = g2d_clean$Gene.symbol,
           Genes2Dis = g2d_ASD,
           Genes = g2d_ASD$Gene.symbol[1:100])
```

## Session Info

```{r}
sessionInfo()
```
